Object-oriented implementations of the MPDATA advection equation solver in~C++,~Python~and~Fortran

### 2013-03-08 paper/doc/paper.tex
suggestions of changes in the part 4 and 5

  One of the aims of this paper is to exemplify the applicability of OOP features of the three considered
    programming languages (or language-library pairs) for representing
    what can be referred to as {\em blackboard abstractions} \citep{Rouson_et_al_2012} within the source code
    of scientific software.
  Presented benchmark tests, although quite simplistic, together with the experience gained 
    from the development of equally-structured codes in three different
    languages provide a basis for discussion on the tradeoffs of the programming language.
  The discussion concerns in principle the development of finite-difference solvers for 
    partial differential equations, but is likely applicable to same extent to
    development of scientific software in general.
  A partly objective and partly subjective summary of pros and cons of C++, Python and Fortran,
    as observed during the development of the presented codes, in presented in the four following subsections.

@@

  It was shown in section~\ref{sec:impl} that C++/Blitz++, Python/NumPy and Fortran
    provide comparable functionalities in terms of matching the blackboard abstractions
    within the program code.
  Taking into account solely the part of code representing particular formul\ae~
    (e.g. listings C.21, P.17, F.20 and equation \ref{eq:antidiff1}) all three
    languages allow to match (or surpass) \LaTeX~in its brevity of formula translation syntax.
  All three languages were shown to be capable of providing mechanisms to compactly represent such abstractions as:
  \begin{itemize}
    \item{loop-free array arithmetics}
    \item{definitions of functions returning array-valued expressions}
    \item{permutations of array indices allowing dimension-independent definitions
      of functions (see e.g. listings C.12 and C.13, P.10 and P.11, F11 and F.12)}
    \item{fractional indexing of arrays corresponding to employment of a staggered grid}
    \item{associating identifiers with slabs of arrays (e.g. for representing 
      a psi[left\_halo]=psi[rght\_edge] assignment)}
  \end{itemize}
  Three issues specific to Fortran were observed which 
    resulted in employment in some places of a more repetitive or cumbersome syntax than in C++ or Python:
  \begin{itemize}
    \item{Fortran does not feature a mechanism allowing to reuse a single piece of code (algorithm)
      with different data types (compare e.g. listings C.6, P.5 and F.4) such as
      templates in C++ and the so-called duck typing in Python;}
    \item{Fortran does not allow function calls to appear on the left hand side
      of assignment (see e.g. how the \prog{tmp} pointers were used as a workaround in the \prog{cyclic\_fill\_halos}
      method in listing F.8)}
    \item{Fortran's built-in array-handling features are not object-oriented,
      and hence not easily extendable via inheritance, but also subject to
      some inherent limitations such as lack of support for arrays of arrays 
      (cf. Sect. \ref{sec:sequence})}
  \end{itemize}
  Interestingly, the limitation in extendability via inheritance was found to
    exist partially in NumPy as well (see footnote \ref{fnt:slice}).
  The lack of a counterpart in Fortran to the C++ template mechanism was identified
    \citep{Cary_et_al_1997}
    as one of the key deficiencies of Fortran when compared with C++ in context 
    of applicability to object-oriented scientific programming.

@@

  Employment of the PyPy drop-in replacement for the standard Python implementation brings 
    Python's performance significantly closer to those of C++ and Fortran, in some
    cases making it the least memory consuming set-up.
  Python has already been the language of choice for scientific software projects having code clarity 
    or ease of use as the first requirement \citep[see e.g.][]{Barnes_and_Jones_2011}.
  PyPy's capability to improve performance of unmodified Python code may actually 
    make Python a favourable choice even if high performance is crucial, especially
    if a performance-to-development cost measure is to be considered.

@@

  Python simplifies also the building and dissemination of software,
    especially if multiple architectures and operating systems are targeted.
  However, CMake\footnote{CMake is a family of open-source, cross-platform
    tools automating building, testing and packaging of C/C++/Fortran software,
    see http://cmake.org/ for details}
    is a tool (or arguably the tool) that allows to efficiently automate 
    builds, tests and packaging of C++ and Fortran programs.

  Python is also definitely easiest to debug among the three languages.
  There exist great debugging tools for C++, but the debugging and development is 
    often hindered by indecipherable compiler messages flooded with lengthy type names
    stemming from employment of templates.
  OOP with Fortran is a relatively new concept, and the support for the Fortran OOP
    features among free and open source compilers, debuggers and other programming
    aids is still not mature.

  A factor contributing to the ease of use of the array-handling 
    facilities is the clarity of rules determining their performance.
  With both Fortran and Python, the memory footprint caused by employment
    of temporary objects in array arithmetics is dependant on compiler optimisations.
  In contrast, Blitz++ ensures temporary-array-free computations by design.
  Similarly, the C++ template mechanism and such features as the C++11
    \prog{constexpr} construct allow the programmer to enforce evaluation
    at compile time\footnote{Useful in context of scientific computing for example to perform
    compile-time dimensional analysis of the physically-meaningful expressions within
    the code with no runtime performance penalty \citet{Schabel_et_al_2008}}.

@@

  The size of the programmers' community of a given language 
    influence the availability of trained personnel, 
    the availability of reusable software components,
    the maturity of compilers and tools, 
    and the volume and coverage of online and printed information resources.
  The OOP features were introduced to Fortran only recently, and have not gained
    wide popularity among users \citep{Worth_2008}\footnote{an anecdotal yet significant
    example being the incomplete support for syntax-highlighting of modern Fortran in Vim and Emacs editors}.
  Fortran is no longer routinely taught at the universities \citep{Kendall_et_al_2008},
    in contrast to C++ and Python.
  An example confirming the arguably ceasing popularity of Fortran in academia 
    is the discontinuation of all but C++ printed editions of the ''Numerical Recipes'' 
    series of Press et al.
  Fortran is a domain-specific language while Python and C++ are general-purpose languages
    with disproportionately larger users' communities.

  Yet still, in the case of C++ and Python it is also the size of the particular
    numerical-library users' community that is of importance.
  Blitz++ is one of several packages that offer high-performance object-oriented
    array manipulation functionality with C++ (and is not necessarily optimal for every
    purpose \citep{Iglberger_et_al_2012}).
  In contrast, the NumPy package became a de facto standard solution for Python.
  Consequently, numerous Python libraries adopted NumPy but
    there are apparently very few C++ libraries offering Blitz++ support out of the box
    (the gnuplot-iostream used in listing C.24 being a much-appreciated counterexample).
  However, Blitz++ allows to instantiate its arrays using preexisting data what
    allows to interface with virtually any library, although requiring additional
    coding.

  The availability and quality of object-oriented 
    libraries differs among the three considered languages.
  The built-in standard libraries of Python and C++ are richer than
    those of Fortran and offer versatile data types, collections of
    algorithms and facilities for interaction with host operating system.
%  External Fortran libraries are mostly limited to scientific applications, 
%    while there exist collections of general-purpose C++ and Python libraries.
%  Interfacing any of the three considered languages from
%    another one does not pose a problem, though.
  In the authors' experience, the small popularity of OOP techniques among
    Fortran users is reflected in the library designs.
  What makes correct use of external libraries more difficult with Fortran
    is the lack of standard exception handling mechanisms in Fortran,
    long and {\em much requested by the numerical community} \citep[][Foreword]{Press_et_al_1996}.

  Finally, the three languages differ as well with regard to availability of 
    mechanisms for leveraging shared-memory parallelisation (e.g. with multi-core processors).
  GCC supports OpenMP with Fortran and C++, as well as the C++11 built-in support for threads.
  The CPython and PyPy implementations of Python do not offer any
    built-in solution for multithreading. 
  However, there are third-party party libraries such as numexpr which can employ multiple cores from
    a single program.
  There exist object-oriented libraries for both C++ and Python to 
    transparently use graphics processing units (GPU) such as Thrust \citep{Thrust}
    and Theano \citep{Bergstra_et_al_2010}.

@@

  Three object-oriented implementations of a prototype solver 
    for the advection equation were introduced.
  The solvers are based on MPDATA - an algorithm of particular applicability
    in geophysical fluid dynamics \citep{Smolarkiewicz_2006}.
  The three implementations follow the same structure but are implemented
    in three different languages (or language/library pairs):
  \begin{itemize}
    \item{C++ with Blitz++,}
    \item{Python with NumPy and}
    \item{Fortran}
  \end{itemize}
  It was the aim of this study to assess the capabilities of these languages
    for object-oriented scientific programming in context of 
    development of finite-difference partial differential equation solvers.
  The emphasis was on the capabilities for reproduction within 
    the program code of the mathematical abstractions used for defining the algorithms
    in journal papers or lecture blackboards.

  Presented programs were developed making use of such recent
    developments as support for C++11 and Fortran 2008 in GCC, and
    the NumPy support in the PyPy implementation of Python.
  The key conclusion is that the considered language/library/compiler
    set-ups offer unprecedented possibilities for using OOP to compactly 
    and faithfully represent within the program source code 
    the mathematical abstractions employed when describing 
    numerical algorithms in literature.
  The readability and brevity of such code may contribute to its 
    auditability, indispensable for credible and reproducible research in computational science 
    \citep{Post_et_al_2005, Merali_et_al_2010, Stodden_et_al_2012}.
  Employment of OOP helps as well to keep complex programs maintainable, avoiding accumulation of the technical/code 
    debt \citep{Buschmann_2011} that besets scientific software in such domains as climate modelling
    \citep{Freeman_et_al_2010}.
  The fact that all considered standards are open and the employed 
    tools implementing them are free and open-source
    is of importance as well \citep{Anel_2011}.

@@

  The OOP design enhances the possibilities to reuse and extend the presented code.
  Development is underway of an object-oriented C++ library featuring concepts presented herein,
    supporting integration in one to three dimensions, handling systems of equations with source terms, 
    providing miscellaneous options of MPDATA and several parallel processing approaches.

>>>>

  One of the aim of this paper is to exemplify OOP features of the three
    programming languages (or language-library pairs)  trying to represent
    what can be referred to as {\em blackboard abstractions} \citep{Rouson_et_al_2012} within the
     scientific software.
  Presented benchmark tests, although quite simplistic, together with the experience gained 
    from the development of equally structured codes in three different
    languages provide a basis for discussion of the tradeoffs of the programming language.
  The discussion concerns in principle the development of finite-difference solvers for 
    partial differential equations, but is likely applicable the scientific software in general.
  A partly objective and partly subjective summary of pros and cons of C++, Python and Fortran,
    as observed during the development of the presented codes, is presented in the four following subsections.

@@

  It was shown in section~\ref{sec:impl} that C++/Blitz++, Python/NumPy and Fortran
    provide comparable functionalities in terms of matching the blackboard abstractions
    within the program code.
  Taking into account solely the part of code representing particular formul\ae~
    (e.g. listings C.21, P.17, F.20 and equation \ref{eq:antidiff1}) all three
    languages allow to match (or surpass) \LaTeX~in its brevity of formula translation syntax.
  All three languages were shown to be capable of providing mechanisms to compactly represent such abstractions as:
  \begin{itemize}
    \item{loop-free array arithmetics;}
    \item{definitions of functions returning array-valued expressions;}
    \item{permutations of array indices allowing dimension-independent definitions
      of functions (see e.g. listings C.12 and C.13, P.10 and P.11, F11 and F.12);}
    \item{fractional indexing of arrays corresponding to employment of a staggered grid;}
    \item{associating identifiers with slabs of arrays (e.g. for representing 
      a psi[left\_halo]=psi[rght\_edge] assignment).} (DOROTA: w zasadzie tylko przyklad rozumiem)
  \end{itemize}
  Three issues specific to Fortran that 
    resulted in employment of a more repetitive or cumbersome syntax than in C++ or Python
    were observed:
  \begin{itemize}
    \item{Fortran does not feature a mechanism allowing to reuse a single piece of code (algorithm)
      with different data types (compare e.g. listings C.6, P.5 and F.4) such as
      templates in C++ and the so-called duck typing in Python;}
    \item{Fortran does not allow function calls to appear on the left hand side
      of assignment (see e.g. how the \prog{tmp} pointers were used as a workaround in the \prog{cyclic\_fill\_halos}
      method in listing F.8)}
    \item{Fortran's built-in array-handling features are not object-oriented,
      and hence not easily extendable via inheritance, but also subject to
      some inherent limitations such as lack of support for arrays of arrays
      (DOROTA: nie podoba mi sie to 'but also', moze wystarczy tylko pierwsza czesc punktu?)
      (cf. Sect. \ref{sec:sequence})}
  \end{itemize}
  Interestingly, the limitation in extendability via inheritance was found to
    exist partially in NumPy as well (see footnote \ref{fnt:slice}).
  The lack of a counterpart in Fortran to the C++ template mechanism was identified
    \citep{Cary_et_al_1997}
    as one of the key deficiencies of Fortran when compared with C++ in context 
    of applicability to object-oriented scientific programming.

@@

  Employment of the PyPy drop-in replacement for the standard Python implementation brings 
    Python's performance significantly closer to those of C++ and Fortran, in some
    cases making it the least memory consuming set-up.
  Python has already been the language of choice for scientific software projects having code clarity 
    or ease of use as the first requirement \citep[see e.g.][]{Barnes_and_Jones_2011}.
  PyPy's capability to improve performance of unmodified Python code may 
    make Python a favourable choice even if high performance is important, especially
    if a performance-to-development cost measure is to be considered.

@@

(DOROTA: zastanawiam sie, czy tego nie wyrzucic)
  Python simplifies also building and shearing of software,
    especially if multiple architectures and operating systems are targeted.
  However, CMake\footnote{CMake is a family of open-source, cross-platform
    tools automating building, testing and packaging of C/C++/Fortran software,
    see http://cmake.org/ for details}
    is a tool (or arguably the tool) that allows to efficiently automate 
    builds, tests and packaging of C++ and Fortran programs.

  Python is also definitely easiest to debug among the three languages.
  There exist great debugging tools for C++, but the debugging and development is 
    often hindered by indecipherable compiler messages.
  OOP with Fortran is a relatively new concept, and the support for the Fortran OOP
    features among free and open source compilers, debuggers and other programming
    aids are still not mature.

  A factor contributing to the ease of use of the array-handling 
    facilities is the clarity of rules determining their performance.
  With both Fortran and Python, the memory footprint caused by employment
    of temporary objects in array arithmetics is dependent on compiler optimisations.
  In contrast, Blitz++ ensures temporary-array-free computations by design.
  Similarly, the C++ template mechanism and such features as the C++11
    \prog{constexpr} construct allow the programmer to enforce evaluation
    at compile time\footnote{Useful in context of scientific computing for example to perform
    compile-time dimensional analysis of the physically-meaningful expressions within
    the code with no runtime performance penalty \citet{Schabel_et_al_2008}}.

@@

  The size of the programmers' community of a given language 
    influence the availability of trained personnel, 
   reusable software components and information resources.
    It also affect the maturity and quality of compilers and tools. 
  Fortran is a domain-specific language while Python and C++ are general-purpose languages
    with disproportionately larger users' communities.
  The OOP features were introduced to Fortran only recently, and have not gained
    wide popularity among users \citep{Worth_2008}\footnote{an anecdotal yet significant
    example being the incomplete support for syntax-highlighting of modern Fortran in Vim and Emacs editors}.
  Fortran is almost no longer taught at the universities \citep{Kendall_et_al_2008},
    in contrast to C++ and Python.
  An example ofdecreasing popularity of Fortran in academia 
    is the discontinuation of Fortran printed editions of the ''Numerical Recipes'' 
    series of Press et al.

  In the case of C++ and Python it is also the size of the particular
    numerical-library users' community that is of importance.
  Blitz++ is one of several packages that offer high-performance object-oriented
    array manipulation functionality with C++ (and is not necessarily optimal for every
    purpose \citep{Iglberger_et_al_2012}).
  In contrast, the NumPy package became a de facto standard solution for Python.
  Consequently, numerous Python libraries adopted NumPy but
    there are apparently very few C++ libraries offering Blitz++ support out of the box
    (the gnuplot-iostream used in listing C.24 being a much-appreciated counterexample).
 ( DOROTA: nie rozumiem nastepnego zdania)
  However, Blitz++ allows to instantiate its arrays using preexisting data what
    allows to interface with virtually any library, although requiring additional
    coding.

(DOROTA: a moze skrocic to i przeniesc do 4.1?)
  The availability and quality of object-oriented 
    libraries differs among the three considered languages.
  The built-in standard libraries of Python and C++ are richer than
    those of Fortran and offer versatile data types, collections of
    algorithms and facilities for interaction with host operating system.
%  External Fortran libraries are mostly limited to scientific applications, 
%    while there exist collections of general-purpose C++ and Python libraries.
%  Interfacing any of the three considered languages from
%    another one does not pose a problem, though.
  In the authors' experience, the small popularity of OOP techniques among
    Fortran users is reflected in the library designs.
  What makes correct use of external libraries more difficult with Fortran
    is the lack of standard exception handling mechanisms in Fortran,
    long and {\em much requested by the numerical community} \citep[][Foreword]{Press_et_al_1996}.

(DOROTA: nastepny akapit zaznaczylabym Fijalowi do uwaznego przejrzenia, 
Ja nie rozumime tej czesci o numexpr itp)
  Finally, the three languages differ as well with regard to availability of 
    mechanisms for leveraging shared-memory parallelisation (e.g. with multi-core processors).
  GCC supports OpenMP with Fortran and C++, as well as the C++11 built-in support for threads.
  The CPython and PyPy implementations of Python do not offer any
    built-in solution for multithreading. 
  However, there are third-party party libraries such as numexpr which can employ multiple cores from
    a single program.
  There exist object-oriented libraries for both C++ and Python to 
    transparently use graphics processing units (GPU) such as Thrust \citep{Thrust}
    and Theano \citep{Bergstra_et_al_2010}.

@@

  Three object-oriented implementations of a prototype solver 
    for the advection equation were introduced.
  The solvers are based on the MPDATA algorithm \citep{Smolarkiewicz_2006}.
  All implementations follow the same structure but are implemented
    in three different language and library pairs:
  \begin{itemize}
    \item{C++ with Blitz++;}
    \item{Python with NumPy;}
    \item{Fortran.}
  \end{itemize}
(DOROTA: przelozylam to i delikatnie zmienilam, ale nadal nie rozuemi czesci 'employed tools...')
  The fact that all considered standards are open and the employed
    tools implementing them are free and open-source
    is certainly a great advantage \citep{Anel_2011}.

  Finite-difference partial differential equation solvers were example used to study 
    the capabilities of these languages
    for object-oriented scientific programming.
  The emphasis was on the capabilities for reproduction 
    of the mathematical abstractions used for explaining the algorithms
    in journal papers within the program code.

  Presented programs were developed making use of such recent
    developments as support for C++11 and Fortran 2008 in GCC, and
    the NumPy support in the PyPy implementation of Python.
  The key conclusion is that all considered language/library/compiler
    set-ups offer possibilities for using OOP to compactly 
     represent the mathematical abstractions within the program  code. 
(OROTA: dwa nastepne zdania bym przeformulowala, ale chce sie upewnic, co wlasciwie chcemy tu powiedziec)
  The readability and brevity of such code may contribute to its 
    auditability, indispensable for credible and reproducible research in computational science 
    \citep{Post_et_al_2005, Merali_et_al_2010, Stodden_et_al_2012}.
  Employment of OOP helps as well to keep complex programs maintainable, avoiding accumulation of the technical/code 
    debt \citep{Buschmann_2011} that besets scientific software in such domains as climate modelling
    \citep{Freeman_et_al_2010}.

@@

(DOROTA: to trzeba zmienic, nie wiem co dokladnie chciales powiedziec.)
  The OOP design enhances the possibilities to reuse and extend the presented code.
  Development is underway of an object-oriented C++ library featuring concepts presented herein,
    supporting integration in one to three dimensions, handling systems of equations with source terms, 
    providing miscellaneous options of MPDATA and several parallel processing approaches.
